---
title: "FOMO-Vignette"
author: "Charles Deng"
package: fomo
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{FOMO-Vignette"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(fomo)
```

# Introduction to FOMO

FOMO is designed to solve mislabeled samples in multi-omic datasets with multiple
samples per individual. The package takes as input sample metadata which includes 
both the expected subject labels of all samples as well as the true grouping of 
samples by genotype. The solver then finds a minimal number of sample relabels 
which aligns subject label groups with genotype groups.

The algorithm is designed to detect and relabel (solve) the following types of
mislabeling events:
1. A simple swap of 2 samples. For example, the sample labeled as sample 1 from 
 individual A (A1) is actually sample 3 from individual B (B3), and the sample 
 labeled B3 should actually be labeled A1.
2. A cycle of N samples. For example, the sample labeled A1 is actually sample B3, 
 B3 is actually C2, and C2 is actually A1.
3. A duplication, when a sample has data generated for both its true sample as 
 well as another planned sample.
4. A deletion, when the true label for a sample doesn't exist in the metadata.
5. Combinations of the above mislabeling events

# Creating a MislabelSolver object

The main object in FOMO is the MislabelSolver. 

### param 1 (required): sample_metadata

The minimal input to create a MislabelSolver a sample_metadata table with three 
columns: one for Sample_ID, one for Subject_ID which groups samples with the 
same subject label, and one for Genotype_Group_ID which groups samples with the 
same true genotype. This last column can be constructed using the pairwise 
identity concordance tool of the researcher's choice.

```{r}

# The minimal inputs required to create a MislabelSolver object
sample_metadata_input <- data.frame(
    # generally comes from metadata records
    Sample_ID = c("A_1", "A_2", "A_3", "B_1", "B_2", "B_3", "C_1", "C_2", "C_3"),
    # generally comes from metadata records
    Subject_ID = c("A", "A", "A", "B", "B", "B", "C", "C", "C"),
    # generally comes from identity concordance tools
    Genotype_Group_ID = c("Group2", "Group1", "Group1", "Group2", "Group2", "Group3", "Group3", "Group1", "Group3")
)

mislabel_solver <- MislabelSolver(sample_metadata_input)

```

### param 2 (optional): genotype_matrix

Alternatively, the genotype group information can be provided as a square, 
symmetric boolean genotype_matrix, with row names and column names coming from 
the Sample_ID column in the sample_metadata and a TRUE indicating that two samples 
have the sample genotype (the diagonal however should all be FALSE). Whenever
a genotype_matrix is provided, the sample_metadata should not include a 
Genotype_Group_ID column.

```{r}

sample_metadata_input <- data.frame(
    # generally comes from metadata records
    Sample_ID = c("A_1", "A_2", "A_3", "B_1", "B_2", "B_3", "C_1", "C_2", "C_3"),
    # generally comes from metadata records
    Subject_ID = c("A", "A", "A", "B", "B", "B", "C", "C", "C")
)

# generally comes from identity concordance tools
genotype_matrix_input <- matrix(FALSE, nrow=9, ncol=9)
rownames(genotype_matrix_input) <- colnames(genotype_matrix_input) <- sample_metadata_input$Sample_ID
genotype_matrix_input[c("A_2", "A_3", "C_2"), c("A_2", "A_3", "C_2")] <- TRUE
genotype_matrix_input[c("A_1", "B_1", "B_2"), c("A_1", "B_1", "B_2")] <- TRUE
genotype_matrix_input[c("B_3", "C_1", "C_3"), c("B_3", "C_1", "C_3")] <- TRUE
diag(genotype_matrix_input) <- FALSE

mislabel_solver <- MislabelSolver(sample_metadata=sample_metadata_input, 
                                  genotype_matrix=genotype_matrix_input)

```

### param 3 (optional): swap_cats

The researcher has the option to provide the argument swap_cats (for swappable
categories), which is a data.frame with two columns: one for Sample_ID which 
should match the Sample_ID column in sample_metadata, and another for SwapCat_ID 
which groups together the Sample_ID(s) which may be swapped for one another. For 
example, if a dataset involves several assay modalities, the modality of each 
sample would be a swappable category provided that sample swaps can only 
occur between samples of the same modality. While this input is optional, the 
additional information provided can help guide and constrain the solver.

```{r}

swap_cats_input <- data.frame(
    # generally comes from metadata records
    Sample_ID = c("A_1", "A_2", "A_3", "B_1", "B_2", "B_3", "C_1", "C_2", "C_3"),
    # generally comes from metadata records
    SwapCat_ID = c("WGS", "RNA", "RNA", "RNA", "RNA", "WGS", "RNA", "WGS", "RNA")
)

mislabel_solver <- MislabelSolver(sample_metadata=sample_metadata_input, 
                                  genotype_matrix=genotype_matrix_input,
                                  swap_cats=swap_cats_input)

```

### param 4 (optional): anchor_samples

If the researcher knows that certain samples are labeled correctly, he/she
can provide that information to the solver through anchor_samples. This is 
represented as a character vector which should be a subset of Sample_ID(s)
from the sample_metadata.

```{r}

anchor_samples_input <- c("C_3")

mislabel_solver <- MislabelSolver(sample_metadata=sample_metadata_input, 
                                  genotype_matrix=genotype_matrix_input,
                                  swap_cats=swap_cats_input,
                                  anchor_samples=anchor_samples_input)

```

### Ghost samples

Not all samples have reliable genotype information, but because they may 
still be involved in mislabeling events with other samples, they should still 
be provided to the solver as "ghost" samples. To specify a ghost sample, set
the corresponding entry in the Genotype_Group_ID column of sample_metadata to NA.
Alternatively if a genotype_matrix was provided instead, remove the row/col 
corresponding to ghost samples from the genotype_matrix.

```{r}

ghost_samples <- c("C_1")
genotype_matrix_input <- genotype_matrix_input[!(rownames(genotype_matrix_input) %in% ghost_samples), 
                                               !(colnames(genotype_matrix_input) %in% ghost_samples)]

mislabel_solver <- MislabelSolver(sample_metadata=sample_metadata_input, 
                                  genotype_matrix=genotype_matrix_input,
                                  swap_cats=swap_cats_input,
                                  anchor_samples=anchor_samples_input)

```

# MislabelSolver functionality

### Plotting

Given an initialized MislabelSolver object, the mislabel network is easily plotted.

Each node is labeled with its Sample_ID
Green nodes indicate that the sample is an anchor sample
Grey nodes indicate that the sample is a ghost samples
Blue edges indicate that the two samples share a common subject label
Grey edges indicate that the two samples sahre a common subject label, but at least one of the two samples is a ghost
Orange edges indicate that the two samples share a common genotype
Green edges indicate that the two samples share both a common genotype and label
Node shapes correspond to swappable categories

```{r}

plot(mislabel_solver)

```

### Solving

FOMO provides four solver functions to find likely relabels: solveMajority, 
solveGlobalSearch, solveLocalSearch, and solveEnsemble. In most cases,
researchers should use the function solveEnsemble, which runs the 
solveGlobalSearch, solveMajority, and solveLocalSearch in a loop.

```{r}

mislabel_solver_solved <- solveEnsemble(mislabel_solver)
## By default, fully concordant cliques are not plotted
## Setting unsolved=FALSE allows fully concordant cliques to be plotted as well
plot(mislabel_solver_solved, unsolved=FALSE)

```

### Plotting corrections

Corrections can be visualized as a directed graph, where a node X pointing to 
node Y indicates that the sample currently labeled X should be relabeled Y

```{r}

plotCorrections(mislabel_solver_solved)

```

### Outputting relabels

Finally, a summary of corrections can be outputted as an Excel spreadsheet

```{r}

summary_list <- writeOutput(mislabel_solver_solved, "~/outputted_corrections")
summary_list$Sample

```